A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-11 tex/codatadp.tex
More work, will revise.

\rn{Complete}
\(
\inferrule{
  v_{\ec{x}} \in FV( \Val ) 
  \quad
  c \in \M
  \quad
  c \not\sim t \sigma \text{ for all } \sigma, t \in \vrange(\Val)
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto c \}
}
\)

%We write $\mathcal{T}(\Val)$ to denote the smallest set containing $v\interp{u}$ for each subterm $u$ of $v$ in the range of $\Val$.
We write $\vrange(\Val)$ to denote the set of $\mu$-terms mapped to by $\Val$.
Since this rule removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.

@@

for which the second premise holds since $\ec{x}$ is the only equivalence class of type $\tau$.

@@

To argue this is the case, notice that either $\tau$ is a recursive type, or contains an infinite subfield of type $\upsilon$.
In the former case, we choose a value $c_1$ whose depth is greater than any term in $\vrange(\Val)$.
In the latter case, we choose a value $c_1$ containing a subterm of type $\upsilon$ not yet occurring in $\Val$.

@@

We now argue the following invariant is maintained by applications of \rn{Complete}:% for all $\ec{x}, \ec{y}$:

@@

%\label{eqn:m-normal} 
%\Val \text{ maps to normal $\mu$-terms only } \\

@@

\Val \ec{x} \sim \Val \ec{y} \text{ if and only if } \ec{x} = \ec{y}

@@

Initially, since \rn{Bisimilar} does not apply to $\Fc$,

@@

and thus~\ref{eqn:m-bij} holds.

@@

Now, we assume $\Val$ satisfies~\ref{eqn:m-bij},
and show that $\Val \sigma$ does as well, where $\sigma := \{ v_{\ec{x}} \mapsto c_1 \}$ is a substitution applied by \rn{Complete}.
By contradiction, assume that $\Val \sigma \ec{y} \sim \Val \sigma \ec{z}$ for some distinct $\ec{y}$ and $\ec{z}$.
Then it must have been the case that $\Val \ec{y}$ and $\Val \ec{z}$ contain a position $p$
where, w.l.o.g., say $\Val \ec{y} \mid_p$ is $v_{\ec{x}}$ and $\Val \ec{z} \mid_p$ is a closed $\mu$-term $c_2$ where $c_1 \sim c_2$.
%Since $\Val$ satisfies invariant~\ref{eqn:m-bij},
%we have that $\Val \ec{w} \sim c_1$ for some $\ec{w}$, 
However, this contradicts the premise of \rn{Complete} that $c_1$ is not bisimilar to any $c_2 \in \mathcal{T}( \Val )$.
Thus, by induction on the number of applications of \rn{Complete}, statement~\ref{eqn:m-bij} holds for $\Val^\ast$.

@@

Now, we let $\M( x )$ be the value in $\M$ that is a renaming of$\Val^\ast \ec{x}$ for each variable $x$ in $\mathcal{T}(\Fc)$,
which are guarenteed to be distinct values due to statement~\ref{eqn:m-bij} and our construction of $\M$.
We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.

@@

If $\tau \in \Data \cup \Codata$,
we first show that $\M( t ) = \Val^\ast \ec{x}$ for all non-selector terms $t$ that occur in equivalence classes $\ec{x}$.
We show this by induction on the size of $\Val^\ast \ec{x}$.
In the case where $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t } = \Val^\ast \ec{ x }$.
In the case where $t$ is a constructor term $C( \vec u )$, 
then $\Val^\ast \ec{x}$ must have been assigned by \rn{Ind} for some constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By the induction hypothesis, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = C( \M( \vec u ) ) = C'( \M( \vec u ' ) ) = \Val^\ast \ec{ x }$.

>>>>

\rn{Complete}
\(
\inferrule{
  v_{\ec{x}} \in FV( \Val ) 
  \quad
  c \in \M
  \quad
  c \not\sim (\mu y. t) \sigma \text{ for all } \sigma, (\mu y. t) \in \mathcal{T}(\Val)
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto c \}
}
\)

We write $\mathcal{T}(\Val)$ to denote the smallest set containing $\nf{(v\interp{u})}$ for each non-variable subterm $u$ of $v$ in the range of $\Val$.
%We write $\vrange(\Val)$ to denote the set of $\mu$-terms mapped to by $\Val$.
Since this rule removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.

@@

for which the third premise holds since $\ec{x}$ is the only equivalence class of type $\tau$.

@@

To argue this is the case, notice that either $\tau \in \Nondata$,
$\tau$ is a recursive type, or contains an infinite subfield of type $\upsilon$.
In the first case, we may choose an arbitrary constant for $c_1$ not yet occurring in $\vrange(\Val)$.
In the second case, we choose a value $c_1$ whose depth is greater than any term in $\vrange(\Val)$.
In the third case, we choose a value $c_1$ containing a subterm of type $\upsilon$ that is not bisimilar to any instance of a term $\mathcal{T}(\Val^\ast)$, 
which is guarenteed to exist by applying this argument inductively.

@@

We further argue that $\Val^\ast$ has the following properties:
%the following invariants are maintained by applications of \rn{Complete}:% for all $\ec{x}, \ec{y}$:

@@

\label{eqn:m-normal} 
\Val^\ast \text{ maps to normal $\mu$-terms only } \\

@@

\Val^\ast \ec{x} \sim \Val^\ast \ec{y} \text{ if and only if } \ec{x} = \ec{y}

@@

We show this by induction on the number of applications of \rn{Complete}.
We additionally show that applications of \rn{Complete} maintain the invariant 
for all subterms $\mu v_{\ec{x}}. t$ occurring in the range of $\Val$ for some $\ec{y}$:
%\begin{eqnarray}
%\label{eqn:m-inv}
%\Val \ec{y} \interp{ \mu v_{\ec{x}}. t } = \Val \ec{x}
%\end{eqnarray}
For the base case, since \rn{Bisimilar} does not apply to $\Fc$,

@@

and thus~\ref{eqn:m-bij} holds for $\Val$.
\rem{todo 1}
%Due to our construction of $\Val$ in step 2 of our calculus, 
%statement~\ref{eqn:m-inv} holds, which together with~\ref{eqn:m-bij} implies~\ref{eqn:m-normal}.

@@

For the inductive case, we assume an arbitrary $\Val_i$ satisfies~\ref{eqn:m-normal} and~\ref{eqn:m-bij},
and show that $\Val_i \sigma_x$ does as well, where $\sigma_x := \{ v_{\ec{x}} \mapsto c_1 \}$ is a substitution applied by \rn{Complete}.
By contradiction, assume that $\Val_i \sigma_x \ec{y} \sim \Val_i \sigma_x \ec{z}$ for some distinct $\ec{y}$ and $\ec{z}$.
Then it must have been the case that $\Val_i \ec{y}$ and $\Val_i \ec{z}$ have a position $p$
where, w.l.o.g., say $\Val_i \ec{y} \mid_p$ is $v_{\ec{x}}$ and $\Val_i \ec{z} \mid_p$ is a term of the form $\mu v_{\ec{w}}. t$ 
and $c_1 \sim (\mu v_{\ec{w}}. t) \sigma_x$.
%Since $\Val_i$ satisfies invariant~\ref{eqn:m-bij},
%we have that $\Val_i \ec{w} \sim c_1$ for some $\ec{w}$, 
However, this contradicts the premise of \rn{Complete} that $c_1$ is not bisimilar to any $t \sigma \in \mathcal{T}( \Val_i )$ for any $\sigma$.
%Thus, by induction on the number of applications of \rn{Complete}, statements~\ref{eqn:m-normal} and~\ref{eqn:m-bij} hold for $\Val^\ast$.
Thus, statements~\ref{eqn:m-normal} and~\ref{eqn:m-bij} hold for $\Val^\ast$.

@@

Now, let $\M( x )$ be the value in $\M$ that is a renaming of$\Val^\ast \ec{x}$ for each variable $x$ in $\mathcal{T}(\Fc)$,
each of which are guarenteed to exist in $\M$ due to statement~\ref{eqn:m-normal} and our construction of $\M$,
and moreover are guarenteed to be distinct in $\M$ due to statement~\ref{eqn:m-bij}.
We first show $\M$ satisfies all equalities in $t_1 \teq t_2 \in \Fc$, where $t_1$ and $t_2$ have type $\tau$.

@@

\rem{revise for co-datatype}
If $\tau \in \Data \cup \Codata$,
we first show that $\M( t ) = \Val^\ast \ec{x}$ for all non-selector terms $t$ that occur in equivalence classes $\ec{x}$.
We show this by induction on the size of $\Val^\ast \ec{x}$.
In the case where $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t } = \Val^\ast \ec{ x }$.
In the case where $t$ is a constructor term $C( \vec u )$, 
then $\Val^\ast \ec{x}$ must have been assigned by \rn{Ind} for some constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By the induction hypothesis, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = C( \M( \vec u ) ) = C'( \M( \vec u ' ) ) = \Val^\ast \ec{ x }$.
